---
title: "Data Import"
output: github_document
---

# Section 1

```{r}
library(tidyverse)
library(readxl)
library(haven)
```

Let's import a dataset.

* read_csv better than read.csv
```{r}
litters_df = 
    read_csv("./data_import_examples/FAS_litters.csv")
```

Look at your data!
```{r}
names(litters_df)
```
* Assessing a variable with a space in it is difficult! 

Update the names in 'litters_df'
```{r}
litters_df = 
    janitor::clean_names(litters_df)
```

Look at the data!
```{r}
litters_df
```

Sometimes skimming data is neat? 
```{r}
skimr::skim(litters_df)

```

Fix the missingness.
```{r}
litters_df = 
    read_csv(file = "data_import_examples/FAS_litters.csv",
        na = c(".", "NA", ""),
    col_types = cols(
      Group = col_factor()
    )
)
```

## Import Pups, missing rows
First 3 lines of FAS_df is missing.
```{r}
pups_df = 
    read_csv("data_import_examples/FAS_pups.csv",
             skip = 3,
               na = c(".", "NA", "")
            )

litters_df = 
    janitor::clean_names(pups_df)

skimr::skim(pups_df)

```


## mlb excel

* CSVs are great, but sometimes you get excel files.

```{r}
mlb_df = 
  read_excel("data_import_examples/mlb11.xlsx")
```
* You can also skip rows, specify sheets, and specify range.

## fotr excel, select range for tables
```{r}
fotr_df = 
    read_excel("data_import_examples/LotR_Words.xlsx", range = "B3:D6")
```

## pulse SAS data
```{r}
pulse_df = read_sas("data_import_examples/public_pulse_data.sas7bdat")

pulse_df = 
  janitor::clean_names(pulse_df)
```
## why do i hate read.csv 

```{r, eval = FALSE}
litters_df_baseR = 
    read.csv("data_import_examples/FAS_litters.csv")
```
* prints our whole dataset, code-formatting isn't useful, no info on rows/columns

## what about data exporting, save data externally
```{r}
write_csv(fotr_df, "data_import_examples/fotr_df.csv")
```
write_csv(df name, "location you want to save at/name of csv.csv")

